---
title: 'Lab #4 Exercises'
subtitle: "Lapse Rates, Clouds, and Water-Vapor Feedback"
author: "put your name here"
date: 'Lab: Mon. Feb.~3 Due: Mon. Feb.~10.'
fontfamily: mathptmx
fontsize: 12pt
output: 
  html_document:
    toc: yes
  github_document:
    toc: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE, include = TRUE, cache = FALSE)

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
data_dir = "_data"
script_dir = "_scripts"

library(pacman)
p_load(tidyverse, jsonlite, httr, magrittr, scales, xml2)

theme_set(theme_bw(base_size = 10))

source(file.path(script_dir, "utils.R"), chdir = T)
source(file.path(script_dir, "format_md.R"), chdir = T)
source(file.path(script_dir, "modtran.R"), chdir = T)
source(file.path(script_dir, "rrtm.R"), chdir = T)
```
# General Instructions

In the past three weeks, we focused on mastering many of the basics of using
R and RMarkdown. For this week's lab, when you write up the answers, I would
like you to think about integrating your R code chunks with your text.

For instance, you can describe what you're going to do to answer the question,
and then for each step, after you describe what you're going to do in that
step, you can include an R code chunk to do what you just described, and then
the subsequent text can either discuss the results of what you just did
or describe what the next step of the analysis will do.

This way, your answer can have several small chunks of R code that build on
each other and follow the flow of your text.

For this lab, you will use the RRTM model, which includes both radiation and
convection.

## Exercise 1: Lapse Rate

Run the RRTM model in its default configuration and then vary the lapse rate
from 0 to 10 K/km. For each value of the lapse rate, adjust the surface
temperature until the earth loses as much heat as it gains (i.e., the value of
_Q_ in the `run_rrtm` model output is zero.)

It will probably be easier to do this with the interactive version of the RRTM
model at <http://climatemodels.uchicago.edu/rrtm/> than with the R interface
`run_rrtm`.

a) Make a tibble containing the values of the lapse rate and the corresponding
   equilibrium surface temperature, and make a plot with lapse rate on the
   horizontal axis and surface temperature on the vertical axis.

b) Describe how the equilibrium surface temperature varies as the lapse rate
   varies.

### Exercise 1 Report:

**Report**: _Use RMarkdown to integrate text, code, graphs, and numbers to 
describe how the equilibrium surface temperature changes as the lapse rate
varies._

# Exercises on Albedo and Clouds

For the following exercises, start off with the RRTM model in its default 
configuration. 
Record the ground temperature.
For each part of this exercise you will do the following:

1. You will adjust the albedo or the clouds.
2. You will compare the visible and longwave radiation going down through the 
   atmosphere to the surface and also the visible and longwave radiation 
   going up from the surface, through the atmosphere, to space.
   
   The results of an RRTM model run have a tibble called `fluxes` with columns
   for `altitude`, `sw_down`, `sw_up`, `lw_down`, `lw_up`, `total_down`, and
   `total_up`, where `sw` means shortwave, `lw` means longwave, and `total`
   is the sum of shortwave plus longwave.
   
   The first row of this tibble is at ground-level and the last row
   is at the top of the atmosphere.
   
   ``` r 
   default_data = run_rrtm()
   fluxes = default_data$fluxes
   surface_fluxes = head(fluxes, 1) # get the first row
   space_fluxes = tail(fluxes, 1)   # get the last row
   ```
   
3. You will adjust the ground temperature until the heat coming in balances
   the heat going out (the model will say, 
   "If the Earth has these properties ... then it loses as much energy as it
   gains."

## Exercise 2: The urban heat island

First, run the RRTM model in its default configuration and note the surface
temperature and the albedo.

```{r RRTM_default}
default = run_rrtm()
albedo_default = default$albedo
T_surface_default = default$T_surface
```

a) Change the surface albedo from "Earth's average" to "Asphalt" and describe
   the changes in the local climate:
   * What is the albedo?
   * Report the changes in shortwave and longwave light absorbed by 
     the surface and going out to space.
   * How much does the total balance of heat change (i.e., how many W/m^2^ does
     the Earth lose or gain)?
   * Adjust the ground temperature until the Earth loses as much energy as it 
     gains. 
   * What is the new surface temperature? 
     How does it compare to the surface temperature in the default 
     configuration?

b) Change the surface albedo to "Concrete". Answer the same questions as in part
   (a).
   
c) In cities, streets and parking lots are usually paved with asphalt.
   Roofs of houses and other buildings are often covered with asphalt shingles 
   or black rubber-like compounds. 

   How would the choice of using asphalt for roads, parking lots, and roofs in
   a large city affect the local climate in the city?
   Would using low-albedo materials, such as concrete for streets and parking 
   lots and light-colored polymers for the roofs of buildings have a benefit 
   for the people living in the city?

### Exercise 2 Report:

**Report**: _Use RMarkdown to integrate text, code, graphs, and numbers to 
describe the effect on the local climate of a city if much of the land is 
covered with asphalt versus if it's covered with concrete. How would this 
affect people living there?_

## Exercise 3: Clouds

First, run the RRTM model in its default configuration and note the surface
temperature and the albedo.

a) Change the low cloud fraction to 0.70 (70%)
   * Report the changes in shortwave and longwave light absorbed by the 
     surface and going out to space.
   * How much does the total balance of heat change 
     (i.e., how many W/m^2^ does the Earth lose or gain)?
   * Adjust the temperature to bring the heat flows back into balance.
   * How much did the temperature change?

b) Repeat part (a), but with the low cloud fraction set to 0 and the high-cloud
   fraction set to 0.20 (20%).
   
c) Use the `plot_heat_flows()` function to plot the heat flows for the 
   low clouds and the high clouds. Describe the changes you see in the upward 
   and downward heat flows (shortwave, longwave, and total) for the two cases.
   Which kind of cloud had the biggest effect on the outgoing radiation?

### Exercise 3 Report:

**Report**: _Use RMarkdown to integrate text, code, graphs, and numbers to 
describe the effect of low clouds and high clouds on heat flow and on the 
surface temperature._
